---
title: "Data cleaning process"
output: 
  html_document: 
    code_folding: hide
    toc: true
    toc_float: true
---

---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)

library(tidyverse)
library(leaflet)

```

## Introduction

This page provides a detailed explanation of the project’s data cleaning and merging process, outlining the logic behind each step, including filtering, feature engineering, and joining.

### Raw data Overview

The data cleaning process involves three distinct NYC Motor Vehicle Collisions datasets. The goal is to produce a unified dataset representing all collisions in the Manhattan borough between January 1, 2021, and October 31, 2025.

The raw data consists of three separate CSV files:

- Crashes: Incident-level details (location, time, summary stats).
- Persons: Details on individuals involved (drivers, passengers, pedestrians).
- Vehicles: Details on the vehicles involved.

### Crash Data Processing

The crash dataset serves as the backbone of the analysis. We perform significant filtering and feature engineering here to establish the scope of the study.

Key Cleaning Steps:

- Scope Filtering: We restrict records strictly to `borough == "MANHATTAN"`.
- Date Parsing: The `crash_date` string is split into numeric components (`month`, `day`, `year`) and converted into a proper Date object.
- Temporal Filtering: We filter for records from 2021 through October 31, 2025.
- Metric Calculation: We aggregate injuries and fatalities into a total `num_casualty` metric.
- Missing Values: Standard placeholders (e.g., "Unspecified", "Unknown") are converted to `NA`.

```{r}
crash_df = read_csv("./data/Motor_Vehicle_Collisions_-_Crashes_20251031.csv", na = c("Unspecified", "Does Not Apply", "Unknown", "NA", ".", "")) |> 
  janitor::clean_names() |> 
  # Keep only Manhattan records
  filter(borough == "MANHATTAN") |>
  
  # Date Decomposition
  separate(
    crash_date,
    into = c("month", "day", "year"),
    sep = "/"
  ) |> 
  # Feature Engineering
  mutate(year = as.numeric(year),
         month = as.numeric(month),
         day = as.numeric(day),
         hour = hour(crash_time),
         num_injured = number_of_pedestrians_injured + number_of_cyclist_injured + number_of_motorist_injured,
         num_killed = number_of_pedestrians_killed + number_of_cyclist_killed + number_of_motorist_killed,
         num_casualty = num_injured + num_killed,
         crash_date_full = make_date(year, month, day), 
         day_of_week = wday(crash_date_full, label = TRUE, abbr = FALSE)
         ) |>
  # Temporal Filtering: keep data from Jan 1, 2021 up to October 31, 2025. 
  filter(
    year %in% 2021:2025 &
    !(year == 2025 & month == 11),
  ) |> 
  relocate(collision_id, year, month, day, hour, day_of_week) |> 
  # Column Selection & Ordering
  select(collision_id, year, month, day, hour, day_of_week, crash_time, zip_code, latitude, longitude,  num_injured, num_killed, num_casualty) |> 
  arrange(year, month, day, hour)
```

### Person Data Processing

The person dataset provides granular detail about the individuals involved.

Key Cleaning Steps:

- Role Standardization: We filter `ped_role` to keep only "Driver", "Passenger", or "Pedestrian", removing ambiguous or administrative roles.
- Renaming: Variables are renamed for clarity (e.g., `ejection` becomes `person_eje`) to prevent confusion after joining.
- Selection: Only demographics (`age`, `sex`) and safety variables are retained.

```{r}
person_df = read_csv("./data/Motor_Vehicle_Collisions_-_Person_20251031.csv", na = c("Unspecified", "Does Not Apply", "Unknown", "NA", ".", "")) |> 
  janitor::clean_names() |> 
  # keep predictors of interests
  select(collision_id, unique_id, vehicle_id, person_type, person_age, ejection, safety_equipment, ped_role, person_sex) |> 
  # keep 3 kinds of main categories of person_role, which account for 99% cases ("registrant" duplicates "driver" most the time)
  filter(ped_role %in% c("Driver", "Passenger", "Pedestrian")) |> 
  # use reasonable names
  rename(person_role = ped_role,
         person_eje = ejection,
         person_safe_equip = safety_equipment,
         person_id = unique_id)
```

### Vehicle Data Processing
The vehicle dataset links specific cars to the crash.

Key Cleaning Steps:

- ID Alignment: Crucially, `unique_id` in this dataset corresponds to `vehicle_id` in the Person dataset. We rename this immediately to facilitate the join key.
- Selection: We retain `vehicle_type`, `year`, and `license_jurisdiction.`

```{r}
vehicle_df = read_csv("./data/Motor_Vehicle_Collisions_-_Vehicles_20251031.csv", na = c("Unspecified", "Does Not Apply", "Unknown", "NA", ".", "")) |> 
  janitor::clean_names() |> 
  # keep predictors of interests
  select(collision_id, unique_id, vehicle_type, vehicle_year, driver_license_status, driver_license_jurisdiction) |> 
  # the unqiue_id in vehicle_df is actually the vehicle_id in person_df, rename it for data joining
  rename(vehicle_id = unique_id)
```


### Data Integration (Joining Strategy)

To create the final analytic dataset, we perform a sequential Left Join strategy starting from crash_df.

- Crash + Person: Joined by `collision_id.`
  - Note: This expands the dataset rows. A single crash with 3 people involved will result in 3 rows.

- Result + Vehicle: Joined by `collision_id` AND `vehicle_id.`
  - Note: This links the specific vehicle characteristics to the driver/passenger in that vehicle.

```{r}

# links every person to their crash details using 'collision_id'.
collision_df = crash_df |>
  left_join(person_df, by = "collision_id") |> 
# links drivers/passengers to their vehicle details using both 'collision_id' and 'vehicle_id'
  left_join(vehicle_df, by = c("collision_id", "vehicle_id")) |> 
  relocate(collision_id, vehicle_id, person_id, year, month, day, hour, day_of_week, crash_time, zip_code, latitude, longitude,  num_injured, num_killed, num_casualty, vehicle_type, vehicle_year, driver_license_status, driver_license_jurisdiction, person_type, person_role, person_age, person_sex, person_eje, person_safe_equip)

write_csv(collision_df, "./data/collision_df.csv")
```
