---
title: "Linear Regression"
date: "2025-12-02"
output:
  html_document: 
    code_folding: hide
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
library(readr)
library(tidyverse)
library(lubridate)
library(rsample)
library(broom)
library(ggmap)
library(viridis)
library(tidymodels)
knitr::opts_chunk$set(echo = TRUE)
```

We use the Manhattan traffic collision records in `collision` as our study data. We first convert the original `crash_time` variable into a standard time format and construct a `datetime` variable, from which we extract date, hour, weekday, month, and a weekend indicator. Based on the cleaned dataset `Data_clean`, we then round `datetime` down to the hour and count the number of collisions in each time interval, obtaining a dataset that contains daily/hourly collision counts and their associated time features (year, month, weekday, weekend, etc.). Using this dataset, we fit time-related linear regression models to investigate the daily patterns of collision events in Manhattan.

```{r}
Data<-read_csv("./Data/collision_df.csv")
```


```{r}
Data_clean <- Data %>%
mutate(
crash_time = as.character(crash_time),
crash_time = ifelse(nchar(crash_time) == 5, paste0(crash_time, ":00"), crash_time),
datetime = ymd_hms(paste(year, month, day, crash_time)),
hour = hour(datetime),
weekday = wday(datetime, label = TRUE, abbr = FALSE),
month = month(datetime, label = TRUE, abbr = FALSE),
weekend = ifelse(weekday %in% c("Saturday","Sunday"), 1, 0),
date = date(datetime)
) %>%
drop_na(datetime)

```




```{r}
hourly_df <- Data_clean %>%
mutate(datetime_hour = floor_date(datetime, "hour")) %>%
count(datetime_hour, name = "num_crash") %>%
mutate(
hour = hour(datetime_hour),
weekday = wday(datetime_hour, label = TRUE, abbr = FALSE, locale = "en_US"),
month = month(datetime_hour, label = TRUE, abbr = FALSE, locale = "en_US"),
year = year(datetime_hour),
weekend = ifelse(weekday %in% c("Saturday","Sunday"), 1, 0)
)

```

## Linear Regression Models
```{r}

ggplot(hourly_df, aes(hour, num_crash)) +
geom_point(alpha = 0.2) +
geom_smooth(method = "loess") +
labs(
title = "Number of accidents by time of day",
x = "Hour",
y = "Number of incidents"
)


hour_weekday_heat <- hourly_df %>%
group_by(weekday, hour) %>%
summarise(mean_crash = mean(num_crash), .groups = "drop")

ggplot(hour_weekday_heat, aes(hour, weekday, fill = mean_crash)) +
geom_tile() +
scale_fill_viridis_c() +
labs(title = "Day × Hour Accident Heat Map", x = "Hour", y = "Day", fill = "Average number of accidents")


```

Ans: 
fit1: base model: hour, weekday, month.
fit2: add the interaction between hour and weekday.
fit3: add the year variable.

Conclusion: Multiple incidents primarily occurred during the morning rush hour starting at 7 a.m. and continuing until 7 p.m., with particularly high concentrations during the morning and evening peak periods. The period from noon to 2 p.m. also showed significant activity, likely due to midday commuting patterns.


## Cross Validation
```{r}
set.seed(123)

cv_splits <- vfold_cv(hourly_df, v = 2)

model_spec <- linear_reg() %>%
set_engine("lm")

wf1 <- workflow() %>%
add_formula(num_crash ~ hour + weekday + month) %>%
add_model(model_spec)

wf2 <- workflow() %>%
add_formula(num_crash ~ hour * weekday + month) %>%
add_model(model_spec)

wf3 <- workflow() %>%
add_formula(num_crash ~ hour + weekday + month + year) %>%
add_model(model_spec)

cv1 <- fit_resamples(wf1, cv_splits, control = control_resamples(save_pred = TRUE))
cv2 <- fit_resamples(wf2, cv_splits, control = control_resamples(save_pred = TRUE))
cv3 <- fit_resamples(wf3, cv_splits, control = control_resamples(save_pred = TRUE))

cv1_metrics <- collect_metrics(cv1)
cv2_metrics <- collect_metrics(cv2)
cv3_metrics <- collect_metrics(cv3)

cv1_metrics
cv2_metrics
cv3_metrics

```

Ans: Among the three models, fit3 (containing hour + weekday + month + year) is the optimal model. Although the improvement is modest, it achieves the best performance on cross-validation metrics. Among the three candidate models, model 3 (hour + weekday + month + year) is selected as the optimal model. This conclusion is based on cross-validation performance. Specifically, model 3 attains the lowest average RMSE and the highest average R² across the 10 folds, indicating slightly better predictive accuracy and explanatory power compared with models 1 and 2. Although the improvement is modest, the performance gains are consistent across folds, suggesting that the inclusion of the year term provides additional predictive value without increasing model variance. Therefore, model 3 is the best-performing model among the three.

## Bootstrap

```{r}
boot_df <- hourly_df %>%
rsample::bootstraps(times = 1000) %>%
mutate(
model = map(splits, ~ lm(num_crash ~ hour + weekday + month, data = analysis(.x))),
coef = map(model, broom::tidy)
) %>%
unnest(coef)

boot_df %>%
filter(term == "hour") %>%
ggplot(aes(x = estimate)) +
geom_density(fill = "steelblue", alpha = 0.5) +
geom_vline(xintercept = 0, color = "red", linetype = "dashed") +
labs(title = "Bootstrap Distribution: Hourly Coefficient", x = "Hourly coefficient estimation", y = "Density")
```

Ans: The hourly coefficient is approximately 0.04–0.055, meaning that for every additional hour, the average number of accidents increases by about 0.04–0.055 incidents. Based on 1,000 bootstrap resamples, the estimated coefficient for hour is highly stable and concentrates within the range 0.04–0.055. This indicates a small but consistent positive relationship between the time of day and the number of crashes. In practical terms, each additional hour is associated with an increase of approximately 0.04–0.055 accidents on average, holding other variables constant. The tight bootstrap distribution suggests that the effect is statistically reliable, even if the magnitude is modest.

## Spatial Heat Map (Incident Hotspots)

```{r}
manhattan_df <- Data_clean %>%
filter(latitude >= 40.7 & latitude <= 40.88,
longitude >= -74.02 & longitude <= -73.93)


ggplot(manhattan_df, aes(x = longitude, y = latitude)) +
stat_density_2d(aes(fill = ..level..), geom = "polygon", contour = TRUE) +
scale_fill_viridis_c(option = "plasma") +
coord_fixed() +
labs(title = "Manhattan Incident Hotspot Map", x = "Longitude", y = "Latitude", fill = "Density")
```














