---
title: "Non_Temporal"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE)
```


```{r}

geo_df <- read.csv("data/Zip Codes.csv") |> 
  janitor::clean_names() |> 
  select(zip_code, county, neighborhood)
  

collision_df <- read.csv("data/collision_df.csv") |> 
  janitor::clean_names() |> 
  drop_na()
```

```{r}
collision_geo_df <- 
collision_df |> 
  left_join(geo_df, by = "zip_code") |> 
  select(-vehicle_year, -driver_license_jurisdiction)
```

```{r}
# Type

collision_geo_df |> 
  count(person_type, name = "number") |> 
  mutate(prop = number / base::sum(number))

# Age

collision_geo_df |> 
  count(person_age, name = "number") |> 
  mutate(prop = number / base::sum(number)) 

# Sex 

collision_geo_df |> 
  count(person_sex, name = "number") |> 
  mutate(prop = number / base::sum(number))

# Role

collision_geo_df |> 
  count(person_role, name = "number") |> 
  mutate(prop = number / base::sum(number))

# Eje

collision_geo_df |> 
  count(person_eje, name = "number") |> 
  mutate(prop = number / base::sum(number))

# Safe equip

collision_geo_df |> 
  count(person_safe_equip, name = "number") |> 
  mutate(prop = number / base::sum(number))

# Safe equip

vehicletype <- collision_geo_df |> 
  count(vehicle_type, name = "number") |> 
  mutate(prop = number / base::sum(number))

```



```{r}
# reset age

collision_reage_df <-  collision_geo_df|>
  filter(person_age >= 0 , person_age <= 140) |> 
  mutate(
    age_group = case_when(
      person_age >= 0 & person_age <= 19 ~ "Children/Teens",
      person_age >= 20 & person_age <= 39 ~ "Young Age",
      person_age >= 40 & person_age <= 64 ~ "Middle Age",
      person_age >= 65 ~ "Old Age"
  ))


collision_reage_df<- collision_reage_df |>
  mutate(
    severity_3level = case_when(
      num_casualty == 0 ~ "low",              
      num_casualty %in% c(1, 2) ~ "middle",    
      num_casualty >= 3 ~ "high",              
      TRUE ~ NA_character_
    )
  )
collision_reage_df |> 
  group_by(age_group) |> 
  summarise(n = n()) |> 
  ggplot(aes(x = forcats::fct_reorder(age_group, n), y = n)) +
  geom_col(fill = "darkblue")+
  labs(title = "Collision Count by Age Group",
       x = "Age Group (Sorted by Count)",
       y = "Total Number of People Involved (Count)")+
  geom_text(aes(label = n), vjust = -0.5, size = 4)


  
```


I set up a filter which only keeps the age: 0 <= person_age <= 140. Then I created a new variable age_group separate to six age group: Pre age are 0 - 19, young age are 20-39, middle age are 40-64, and old age begins around 65.
After the graph we can see the most of the crash are happened to the Young age, secondly are the Middle age. 




```{r}
collision_reage_df |>
  filter(!is.na(age_group)) |>
  group_by(age_group, severity_3level) |>
  summarise(n = n(), .groups = 'drop') |>
  ungroup() |>
  group_by(age_group) |>
  mutate(prop = n / sum(n)) |>
  mutate(
    age_group = factor(
      age_group,
      levels = c("Children/Teens", "Young Age", "Middle Age", "Old Age")
    )
  ) |>
  ggplot(aes(x = severity_3level, y = age_group, fill = prop)) +
  geom_tile(color = "darkblue", linewidth = 0.5) +
  scale_fill_viridis_c(
    option = "magma",
    labels = scales::percent,
    direction = -1
  ) +
  geom_text(aes(label = scales::percent(prop, accuracy = 0.1)), color = "grey", size = 3) +
  labs(
    title = "Proportional Heatmap of Collision Severity by Age Group",
    x = "Collision Severity",
    y = "Age Group",
    fill = "Proportion"
  ) +
  scale_x_discrete(limits = c("low", "middle", "high")) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

```




```{r}
# Person Ejected and Person_Safe_Equip regroup

collision_popcar_df <- collision_reage_df |>
  filter(vehicle_type == "Sedan" | vehicle_type == "Station Wagon/Sport Utility Vehicle"| vehicle_type == "Taxi") |> 
  mutate(
    ejected_binary = case_when(
      person_eje %in% c("Ejected", "Partially Ejected", "Trapped") ~1,
      person_eje == "Not Ejected" ~ 0,
    )
  ) |>
  mutate(
    seatbelt_used = case_when(
      person_safe_equip %in% c(
        "Lap Belt", 
        "Harness", 
        "Lap Belt & Harness",
        "Air Bag Deployed/Lap Belt", 
        "Air Bag Deployed/Lap Belt/Harness",
        "Child Restraint Only",
        "Air Bag Deployed/Child Restraint"
      ) ~ 1, 
      person_safe_equip %in% c("None") ~ 0
    )
  ) |> 
  drop_na() 

collision_popcar_df|> 
  group_by(seatbelt_used) |> 
  summarise(n = n()) |> 
  knitr::kable()

```


```{r}
logistic_fit <- 
  glm(ejected_binary ~ seatbelt_used, 
  data = collision_popcar_df, 
  family = binomial()
)

broom::tidy(logistic_fit, exponentiate = TRUE)
```





```{r}
collision_popcar_df <- collision_popcar_df |> 
  select(vehicle_id, person_id, num_injured, num_killed, num_casualty, vehicle_type, driver_license_status, person_role, person_age, person_sex,age_group,seatbelt_used, ejected_binary)


```















```{r}
 
collision_geo_df |> 
  count(neighborhood, name = "number") |>
  mutate(
    prop = number / base::sum(number)
  ) |> 
  arrange(desc(number)) |> 
  knitr::kable(digits = 2)


# Visualizes the count of collisions by Neighborhood, separated by Person Role (the facet), and grouped within the bars by Person Sex.

collision_geo_df |> 
  drop_na(person_role, person_sex) |> 
  ggplot(aes(x = fct_reorder(neighborhood, neighborhood, .fun = length), fill = person_sex))+
  geom_bar(position = "dodge")+
  facet_wrap(~ person_role, scales = "free_y") +
  labs(title = "Collision Counts by Neighborhood, Grouped by Sex",
       x = "Neighborhood (Sorted by Count)",
       y = "Number of Collisions") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```

