---
title: "Non_Temporal"
output: html_document
---

```{r}
library(tidyverse)
```


```{r}

geo_df <- read.csv("data/Zip Codes.csv") |> 
  janitor::clean_names() |> 
  select(zip_code, county, neighborhood)
  

collision_df <- read.csv("data/collision_df.csv") |> 
  janitor::clean_names()
```

```{r}
collision_geo_df <- 
collision_df |> 
  left_join(geo_df, by = "zip_code") |> 
  select(-vehicle_year, -driver_license_jurisdiction) |> 
  drop_na()
```

```{r}
# Type

collision_geo_df |> 
  count(person_type, name = "number") |> 
  mutate(prop = number / base::sum(number))

# Age

collision_geo_df |> 
  count(person_age, name = "number") |> 
  mutate(prop = number / base::sum(number)) 

# Sex 

collision_geo_df |> 
  count(person_sex, name = "number") |> 
  mutate(prop = number / base::sum(number))

# Role

collision_geo_df |> 
  count(person_role, name = "number") |> 
  mutate(prop = number / base::sum(number))

# Eje

collision_geo_df |> 
  count(person_eje, name = "number") |> 
  mutate(prop = number / base::sum(number))

# Safe equip

collision_geo_df |> 
  count(person_safe_equip, name = "number") |> 
  mutate(prop = number / base::sum(number))
```



```{r}
# reset age

collision_reage_df <-  collision_geo_df|>
  filter(person_age >= 0 , person_age <= 140) |> 
  mutate(
    age_group = case_when(
      person_age >= 0 & person_age <= 19 ~ "Pre young",
      person_age >= 20 & person_age <= 39 ~ "Young Age",
      person_age >= 40 & person_age <= 64 ~ "Middle Age",
      person_age >= 65 ~ "Old Age"
  ))

collision_reage_df |> 
  group_by(age_group) |> 
  summarise(n = n()) |> 
  ggplot(aes(x = forcats::fct_reorder(age_group, n), y = n)) +
  geom_col(fill = "darkblue")+
  labs(title = "Collision Count by Age Group",
       x = "Age Group (Sorted by Count)",
       y = "Total Number of People Involved (Count)")+
  geom_text(aes(label = n), vjust = -0.5, size = 4)
  
```


I set up a filter which only keeps the age: 0 <= person_age <= 140. Then I created a new variable age_group separate to six age group: Pre age are 0 - 19, young age are 20-39, middle age are 40-64, and old age begins around 65.
After the graph we can see the most of the crash are happened to the Young age, secondly are the Middle age. 



```{r}
 
collision_geo_df |> 
  count(neighborhood, name = "number") |>
  mutate(
    prop = number / base::sum(number)
  ) |> 
  arrange(desc(number)) |> 
  knitr::kable(digits = 2)


# Visualizes the count of collisions by Neighborhood, separated by Person Role (the facet), and grouped within the bars by Person Sex.

collision_geo_df |> 
  drop_na(person_role, person_sex) |> 
  ggplot(aes(x = fct_reorder(neighborhood, neighborhood, .fun = length), fill = person_sex))+
  geom_bar(position = "dodge")+
  facet_wrap(~ person_role, scales = "free_y") +
  labs(title = "Collision Counts by Neighborhood, Grouped by Sex",
       x = "Neighborhood (Sorted by Count)",
       y = "Number of Collisions") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```


