---
title: "Project Proposal: Motor Vehicle Collisions in Manhattan"
author: "Liliang Su"
date: "2025-11-05"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Group Members

```{r, echo = FALSE}
tibble(
  Name = c("Liliang Su", "Anny Ding", "Haohua Chen", "Chenhao Zheng", "Duan Yan"),
  UNI = c("ls4236", "cd3591", "hc3654", "cz2955", "yd2865")
) |> 
  knitr::kable()
```


## 2. Tentative Project Title

Motor Vehicle Collisions in Manhattan: Identifying Peak Risk Times and Temporal Hotspots for Traffic Incidents


## 3. Motivation for this Project

- The preliminary data structure indicates a rich dataset of traffic incidents with precise temporal markers (`year`, `month`, `day`, `crash_time`). The primary motivation is to leverage this granularity to understand when incidents are most likely to occur. 

- By identifying high-risk time windows (e.g., specific hours, days of the week, or seasonal patterns), we can provide actionable insights for traffic management, public safety messaging, and resource allocation to potentially reduce future crashes. The goal is to move beyond simple incident counts and uncover meaningful, predictive temporal patterns.


## 4. Intended Final Products

- **Comprehensive R Markdown Report**: A fully reproducible report detailing data cleaning, robust temporal aggregation, statistical findings, and interpretation of peak risk times.

- **Interactive Visualization**: A web application displaying key visualizations that allow a user (e.g., a city planner) to dynamically filter incident frequency by year, month, day of the week, and hour of the day.

- **Develop Predictive Tools**: Create a model that can forecast the likelihood of a crash at a given time, enabling law enforcement and emergency services to optimize resource deployment and implement targeted prevention campaigns.


## 5. Anticipated Data Sources

- **Primary Data source**: 

- 3 NYC Motor Vehicle Collisions related dataset (Crashes, Vehicles, Person) collected from [NYC Open Data](https://data.cityofnewyork.us/browse?sortBy=relevance&pageSize=20&limitTo=datasets&q=NYC+Traffic+Accidents&page=1).

- We only focus on Manhattan observations in recent 5 years (01/01/2021 -- 10/31/2025)

- **Anticipated Auxilary Data**:
  - Weather Data in Manhattan
  - Personal Income Data in Manhattan
  
## 6. Planned Analyses / Visualizations / Coding Challenges

- Planned Analyses / Visualizations

  - **Temporal Heatmaps**: Visualize crash counts by hour of day vs. day   of week to pinpoint rush-hour risk.
  
  - **Seasonal Trends**: Time-series plots of monthly crash totals to   identify seasonal peaks.
  
  - **Spatial Clustering**: Generating kernel density maps (heatmaps) to   highlight high-density crash locations (hotspots).
  
- Coding Challenges

  - **Data Wrangling**: Converting the separate `year`, `month`, `day`, and `crash_time` columns into a single, cohesive datetime object for efficient analysis, and generating features like "Day of Week" and "Is Weekend".
  
  - **Integration Different Data Source**: Successfully integrate multiple datasets to produce effective visualizations.
  
  - **Time-Series Modeling** (Tentative): Implementing and validating a time-series model (e.g., ARIMA or a generalized linear model) to predict future crash counts.
  
## 6. Planned Timeline

```{r, echo = FALSE}
tibble(
  Week = c("Week1", "Week2", "Week3", "Week4" ),
  Focus_Area = c("Data Preparation and tidy up", "Temporal Analysis and visualization", "Spatial and Multivariate Analysis", "Synthesis and Final Repor"),
  Key_Deliverables = c("Convert year/month/day/time into a single datetime object and clean the dataset", 
                       "Heatmaps (Hour vs. Day) and time-series plots to identify high-risk time windows and seasonal patterns. Successfully integrate the first external dataset (e.g., weather).", 
                       "Generate spatial crash hotspot maps (if location data is present). Explanatory Visualizations: Analyze the relationship between crash frequency/severity and external factors (e.g., weather, road type).", 
                       "Synthesize all findings and visualizations. Conclusions: Draft clear, actionable policy and safety recommendations. Submit the final, polished R Markdown report.")
  
) |> 
  knitr::kable()
```
